---
title: "eda"
output: html_document
date: "2025-10-02"
---


```{r setup, include=FALSE}
library(tidyverse)
library(p8105.datasets)
```


```{r - default options}
knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```


Import the weather data

```{r}
data("weather_df")

weather_df = 
  weather_df |> 
  mutate(month = floor_date(date, unit = "month"))
      # floor_date rounds down to first of month
      #lubridate has helpful date functions

```


make plots
```{r}
weather_df |> 
  ggplot(aes(x=prcp))+
  geom_histogram()

  # see that right skewed
```


check on extreme values

```{r}
weather_df |> 
  filter(prcp > 1000)
```

Look at data again

```{r}
weather_df |> 
  filter(tmax >= 20, tmax <= 30) |> 
  ggplot(aes(x=tmin, y = tmax, color = name, shape = name)) +
  geom_point(alpha = 0.75)

    # see that he washington data points are kind of all over
```


group_by

```{r}
weather_df |> 
  group_by(name, month)
    # can group by multiple variables
    # group by doesnt cange anyothing or add any variables
```

group and count things

```{r}
weather_df |>
  group_by(name) |>
  summarize(n = n())

    # can see how many obs we have for each group
    # first we gorup by name, then we summarize the obs we have
```

```{r}
weather_df |>
  group_by(month) |>
  summarize(n = n())

    # seeing number of obs by month
    # group by name and month shows us how many days in each month
```

```{r}
weather_df |>
  group_by(month) |>
  summarize(n = n_distinct(date))

  # shows number of distinct dates
```

you can count directly
```{r}
weather_df |> 
  count(name)

  # same as above
```


```{r}
weather_df |> 
  group_by(month) |> 
  summarize(
    mean_tmax = mean(tmax, na.rm = T),
    median_tmin = median(tmin, na.rm = T),
    sd_prcp = sd(prcp, na.rm = T)
  )

    # but these avgs are across all the different locations which isn't very helpful
```


```{r}
weather_df |> 
  group_by(name, month) |> 
  summarize(
    mean_tmax = mean(tmax, na.rm = T),
    median_tmin = median(tmin, na.rm = T),
    sd_prcp = sd(prcp, na.rm = T)
  )
```

This is still a dataframe
```{r}
weather_df |> 
  group_by(name, month) |> 
  summarize(
    mean_tmax = mean(tmax, na.rm = T)
  ) |> 
  ggplot(aes(x = month, y = mean_tmax, color = name)) +
  geom_point()+
  geom_line()
```

sometimes format results more nicely

```{r}
weather_df |> 
  group_by(name, month) |> 
  summarize(
    mean_tmax = mean(tmax, na.rm = T)
  ) |>
  pivot_wider(
    names_from = name, 
    values_from = mean_tmax
  ) |> 
  knitr::kable(digits = 2)
    #knitr creates a nice looking table
```

## mutate with groups

```{r}
weather_df |> 
  group_by(name) |> 
  mutate(
    mean_tmax = mean(tmax, na.rm = T),
    center_tmax = tmax-mean_tmax
  ) |> 
  ggplot(aes(x = date, y = center_tmax, color = name)) +
  geom_point()
# for each weather station, what is avg, remove mean, then plot result
# centering the values so the center is 0
```

look for cold daays

```{r}
weather_df |> 
  group_by(name, month) |> 
  mutate(temp_rank = desc(min_rank(tmax))) |> 
  filter(temp_rank < 2)

    # showing the coldest day bc temp_rank <2
```

WHat about lags
Does my current value depend on previous values
did yesterdays stress level predict today's sleep quality?
```{r}
weather_df |> 
  group_by(name) |> 
  mutate(lagged_tmax = lag(tmax))
 # lag is the difference ebtween today and yesterdays day
  #the group_by is added here bc...
```

use the variables you create
```{r}
weather_df |> 
  group_by(name) |> 
  mutate(
    temp_change = tmax - lag(tmax)
  ) |> 
  summaraize(
    sd_tmax_change = sd(temp_change, na.rm = T),
    tmax_change_max = max(temp_change, na.rm = T)
  )

# the biggest change 
```

which day did the max change happen
```{r}
weather_df |> 
  group_by(name) |> 
  mutate(
    temp_change = tmax - lag(tmax),
    change_rank = min_rank(desc(temp_change))
  ) |> 
  filter(
    change_rank < 2
  )
```

