---
title: "viz_1"
output: github_document
date: "2025-09-25"
---

```{r setup, include=FALSE}
library(tidyverse)
library(patchwork)
library(p8105.datasets)
```

Import the weather data

```{r}
data("weather_df")
```

Let's make out basic scatterplot

```{r}
weather_df |> 
  ggplot(aes(x = tmin, y = tmax))+
  geom_point(aes(color = name), alpha = 0.5) +
  labs(
    x = " Min daily temp",
    y = " max daily temp",
    title = "Temp scatterplot",
    caption = "Data from NOAA",
    color = "Location" # color for different location
  )
```


## Scales

scale for x and y axis

```{r}
ggplot_temperature =
  weather_df |> 
  # filter(tmax > 10, tmax <30) # another way to zoom in
  ggplot(aes(x = tmin, y = tmax))+
  geom_point(aes(color = name), alpha = 0.5) +
  labs(
    x = " Min daily temp",
    y = " max daily temp",
    title = "Temp scatterplot",
    caption = "Data from NOAA",
    color = "Location" # color for different location
  ) +
  scale_x_continuous(
    breaks = c(-20, 0 , 25),
    labels = c("-20C", "0", "25")
    ) + #X is continuous so using this argument
  scale_y_continuous(
    trans = "sqrt",
    limits = c(10,30) # zooming in to specific y axis scale
  ) +
 # scale_color_hue (h= c(100,300)) # changes color palette. but actually use viridis package
  viridis::scale_color_viridis(
    discrete = TRUE
  ) 
```

## Themes

change everything going on with the plot

Making my base plot
  # theme_bw, theme_minimal

```{r}
ggplot_temperature +
  #theme_minimal() +
  theme(legend.position = "bottom")
  
```


```{r}
ggsave("weather_scatterplot.png", ggplot_temperature) 
```


## Adding data in geoms

```{r}
central_park_df = 
  weather_df |> 
  filter(name == "CentralPark_NY")

molokai_df = 
  weather_df |> 
  filter(name == "Molokai_HI")

ggplot(data = molokai_df, aes(x=date, y=tmax, color = name)) +
  geom_point() +
  geom_line(data = central_park_df)  #this is helpful if you want to add in monthly avg, you would create a different df with the avg
  
```


## `patchwork`

make 3 plots and combine using patchwork

first is scatterplot between t max and t min
second is density plot of recip
thrid shows seasonality and groupings of 

```{r}
ggp_tmax_tmin =
  weather_df |> 
  ggplot(aes(x=tmax, y = tmin, color = name))+
  geom_point(alpha = 0.5)+
  theme(legend.position = "none")

ggp_prec_density = 
  weather_df |>
  filter(prcp > 0) |> 
  ggplot(aes(x = prcp, fill = name)) +
    geom_density(alpha = 0.5)+ 
  theme(legend.position = "none")


(ggp_tmax_tmin + ggp_prec_density)
  
```


## Data manipulation

temp violin plots

```{r}
weather_df |>
  mutate(name = forcats::fct_relevel(name, c("Molokai_HI", "CentralPark_NY", "Waterhole_WA"))) |> # specifying the orde we wantt o see the categoreis in in graph
  ggplot(aes(x = name, y = tmax)) + 
  geom_violin(aes(fill = name), color = "blue", alpha = .5) + 
  theme(legend.position = "bottom")
```

name in order of avg value for each of the groups

```{r}
weather_df |>
  mutate(name = forcats::fct_reorder(name, tmax)) |> # this is helpful if you have a lot of categories an you want them in order by avg avlue 
  ggplot(aes(x = name, y = tmax)) + 
  geom_violin(aes(fill = name), color = "blue", alpha = .5) + 
  theme(legend.position = "bottom")
```

what about data tidiness

does distribution of the depression index change over time . 

telling ggplot to make viz where x = study visit, y = score. box plot. but our data aren't structured in a way to make this box plot


```{r}
pulse_df = 
  haven::read_sas("data/public_pulse_data.sas7bdat") |> 
  janitor::clean_names() |> 
  pivot_longer(
   bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    names_prefix = "bdi_score_",
    values_to = "bdi") |>
  mutate(visit = fct_inorder(visit)) #fct_inorder adds the variables in the way they show up. baseline show sup first so thats why it is first now
```

```{r}
ggplot(pulse_df, aes(x = visit, y = bdi)) + 
  geom_boxplot()
```


thinking thorught he pups figure

x = dose bc we see con, low, mod
y = days until outcome
facet - the two rows are day of tx
facet - the four columns of specific outcomes

variables we need: pup, dose, tx, outcome, postnatal day

make a plot for the FAS study

```{r}
pups_df = 
  read_csv("data/FAS_pups.csv", na = c("NA", ".", ""), skip = 3) |> 
janitor::clean_names() |> 
  mutate(
    sex = 
      case_match(
        sex, 
        1 ~ "male", 
        2 ~ "female"))


litters_df = 
  read_csv("./data/FAS_litters.csv", na = c("NA", ".", "")) |>
  janitor::clean_names() |>
  separate(group, into = c("dose", "tx_day"), sep = 3)

# group has 2 pieces of info I need: dose and tx day
```


```{r}
fas_data = left_join(pups_df, litters_df, by = "litter_number") 

fas_data |> 
  select(pd_ears:tx_day) |> 
  pivot_longer(
    pd_ears:pd_walk,
    names_to = "outcome",
    names_prefix = "pd_",
    values_to = "pn_day"
  ) |> 
  mutate(outcome = fct_reorder(outcome, pn_day)) |> 
  drop_na() |> 
  ggplot(
    aes(x = dose, y = pn_day) )+
    geom_violin()+
    facet_grid(tx_day~outcome)
```





